---
title: "Chapter 3"
subtitle: "Distributions"
output: 
  html_document:
    number_sections: yes
---

```{r load_packages3, message = FALSE, warning=FALSE}
# Packages required for Chapter 3
library(gridExtra)  
library(knitr) 
library(kableExtra)
library(tidyverse)
```


# Discrete distributions

## Binary (Bernoulli)

## Binomial


```{r, multBin, fig.align="center",out.width="60%",fig.cap= '(ref:multBin)', echo=FALSE, warning=FALSE, message=FALSE}
#binomialPlots
plotBinomial=function(n,p){
  y1 <- 0:n  # possible values
  prob <- dbinom(y1,n,p)  # P(Y=y)
  BBGdf <- data.frame(y1,prob)
  ggplot(data = BBGdf, aes(x = y1, xend = y1, y = 0, yend = prob)) + 
    geom_segment() + 
    xlab("number of successes") + ylab("probability") + labs(title=paste("n = ", n, " p = ", p))
}
Binom1 <- plotBinomial(10,.25) + xlim(0, 10) + scale_y_continuous(breaks = c(0.00, 0.05, 0.10, 0.15, 0.20, 0.25))
Binom2 <- plotBinomial(20,.2) + xlim(0, 15)
Binom3 <- plotBinomial(10,.5) + xlim(0, 10)
Binom4 <- plotBinomial(50,.2) + xlim(0, 25)

grid.arrange(Binom1,Binom2,Binom3,Binom4,ncol=2)
```

Example density function

```{r}
dbinom(2, size = 10, prob = .25)
```

## Geometric


```{r, multGeo, fig.align="center",out.width="60%",fig.cap= '(ref:multGeo)', echo=FALSE, warning=FALSE, message=FALSE}
#geometricPlots
plotGeo <- function(p){
yg = 0:10000 # possible values
gd <- data.frame(x=rgeom(yg, p))  # generate random deviates
#breaks <- pretty(range(gd$x), n = nclass.FD(gd$x), min.n = 1)  # pretty binning
#bwidth <- .5              #breaks[2] - breaks[1]
ggplot(gd, aes(x = x)) + geom_histogram(aes(y=..count../sum(..count..)), binwidth = .25) +
  xlab("number of failures") + ylab("probability") + labs(title=paste("Geometric p = ", p)) + xlim(-1,16)
}

Geo1 <- plotGeo(0.3)
Geo2 <- plotGeo(0.5)
Geo3 <- plotGeo(0.7)
grid.arrange(Geo1,Geo2, Geo3, ncol=1)
```

Example density function

```{r geomEx}
dgeom(2, prob = 1/6)
```

## Negative Binomial


```{r, multNBinom, fig.align="center",out.width="60%",fig.cap= '(ref:multNBinom)', echo=FALSE, warning=FALSE, message=FALSE}
#negativeBinomialPlots
plotNBinom <- function(p, r){
  ynb <- 0:10000
  nbd <- tibble(x = rnbinom(ynb, r, p))
  #breaks <- pretty(range(nbd$x), n = nclass.FD(nbd$x), min.n = 1)  # pretty binning
  #bwidth <- breaks[2] - breaks[1]
  ggplot(nbd, aes(x = x)) +
    geom_histogram(aes(y=..count../sum(..count..)), binwidth = .25) +
    labs(title = paste("p = ", p, ", r = ", r),
         x = "number of failures",
         y = "probability") +
    xlim(-1,30)
}

NBin1 <- plotNBinom(0.35, 3)
NBin2 <- plotNBinom(0.35, 5)
NBin3 <- plotNBinom(0.70, 5)
grid.arrange(NBin1, NBin2, NBin3, ncol = 1)
```

Example density function

```{r}
sum(dnbinom(0:2, size = 10, prob = .9))  
```

## Hypergeometric


```{r multHyper, fig.align="center",out.width="60%",fig.cap= '(ref:multHyper)', echo=FALSE, warning=FALSE, message=FALSE}
#hyperGeometric Plots
plotHyper <- function(m, N, k){
  yp <- 0:10000 
  hd <- data.frame(x = rhyper(yp, m, N-m, k))
  ggplot(hd, aes(x = x)) + geom_histogram(aes(y=..count../sum(..count..)), binwidth = .25) +
  xlab("number of successes") + ylab("probability") + labs(title=paste("m = ", m, ",", "N = ", N, ",", "n = ", k))
}

Hyp1 <- plotHyper(25,1000,100) + xlim(-1,25) + scale_y_continuous(breaks = c(0.00, 0.05, 0.10, 0.15, 0.20, 0.25))
Hyp2 <- plotHyper(200,1000,500) + xlim(80,120)
Hyp3 <- plotHyper(100,1000,100) + xlim(-1,25)
Hyp4 <- plotHyper(200,1000,950) + xlim(160,200)
grid.arrange(Hyp1, Hyp2, Hyp3, Hyp4, ncol=2)
```

Example density function

```{r}
dhyper(4, m = 4, n = 48, k = 10)
```

## Poisson


```{r, multPois, fig.align="center",out.width="60%",fig.cap= '(ref:multPois)', echo=FALSE, warning=FALSE, message=FALSE}
#poissonPlots
plotPois <- function(lam){
yp = 0:10000 # possible values
pd <- data.frame(x=rpois(yp, lam))  # generate random deviates
breaks <- pretty(range(pd$x), n = nclass.FD(pd$x), min.n = 1)  # pretty binning
#bwidth <- breaks[2] - breaks[1]
ggplot(pd, aes(x = x)) + geom_histogram(aes(y=..count../sum(..count..)), binwidth = .25) +
  xlab("number of events") + ylab("probability") + 
  labs(title=paste("Poisson lambda = ", lam)) + xlim(-1,13)
}

Pois1 <- plotPois(0.5)
Pois2 <- plotPois(1)
Pois3 <- plotPois(5) + scale_y_continuous(breaks = c(0, 0.1))
grid.arrange(Pois1,Pois2,Pois3,ncol=1)
```

Example density function

```{r}
dpois(0:10, lambda = .5)
```


# Continuous distributions

## Exponential


```{r, multExp, fig.align="center",out.width="60%",fig.cap= '(ref:multExp)', echo=FALSE, warning=FALSE, message=FALSE}
#exponentialPlots
x=seq(0,4,by=0.01)  # possible values
probex1 <- dexp(x,.5)  # P(Y=y)
probex2 <- dexp(x,1)
probex3 <- dexp(x,5)
Expdf <- tibble(x,probex1, probex2, probex3) %>%
  rename(x = x,
         `0.5` = probex1,
         `1` = probex2,
         `5` = probex3) %>%
  gather(2:4, key = "Lambda", value = "value") %>%
  mutate(Lambda = factor(Lambda, levels = c("0.5", "1", "5")))
ggplot(data = Expdf, aes(x = x, y = value, color = Lambda)) +
  geom_line(aes(linetype = Lambda)) +
  xlab("values") + ylab("density") + 
  labs(title = "Exponential Distributions") + 
  xlim(0,4) + ylim(0,3)
```

Example probability function

```{r}
pexp(10, rate = 1/6)
```

## Gamma


```{r, multGamma, fig.align="center",out.width="60%",fig.cap= '(ref:multGamma)', echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 7, by = 0.01)
`r = 1, lambda = 1` <- dgamma(x, 1, rate = 1)
`r = 2, lambda = 1` <- dgamma(x, 2, rate = 1) 
`r = 5, lambda = 5` <- dgamma(x, 5, rate = 5)
`r = 5, lambda = 7` <- dgamma(x, 5, rate = 7)

gammaDf <- tibble(x, `r = 1, lambda = 1`, `r = 2, lambda = 1`, `r = 5, lambda = 5`, `r = 5, lambda = 7`) %>%
  gather(2:5, key = "Distribution", value = "value") %>%
  mutate(Distribution = factor(Distribution, 
                               levels = c("r = 2, lambda = 1", 
                                          "r = 1, lambda = 1", 
                                          "r = 5, lambda = 5", 
                                          "r = 5, lambda = 7")))

ggplot(data = gammaDf, aes(x = x, y = value, 
                           color = Distribution)) +
  geom_line(aes(linetype = Distribution)) +
  xlab("values") + ylab("density") + 
  labs(title = "Gamma Distributions") +
  theme(legend.title = element_blank())
```


```{r}
pgamma(3, shape = 5, rate = 2)
```

## Normal (Gaussian)

```{r, multNorm, fig.align="center",out.width="60%",fig.cap= '(ref:multNorm)', echo=FALSE, warning=FALSE, message=FALSE}
#normalPlots
x <- seq(-10, 20, by = 0.01)
norm1 <- dnorm(x, -5, 2)
norm2 <- dnorm(x, 0 , 1)
norm3 <- dnorm(x, 5, 5)
norm4 <- dnorm(x, 0, 3)

normDf <- tibble(x, norm1, norm2, norm3, norm4) %>%
  rename(`N(-5, 2)` = norm1,
         `N(0, 1)` = norm2,
         `N(10, 5)` = norm3,
         `N(0, 3)` = norm4) %>%
  gather(2:5, key = "Distribution", value = "value") %>%
  mutate(Distribution = factor(Distribution, 
    levels = c("N(10, 5)","N(0, 3)", "N(0, 1)", "N(-5, 2)")))

ggplot(data = normDf, 
       aes(x = x, y = value, color = Distribution)) +
  geom_line(aes(linetype = Distribution)) +
  xlab("values") + ylab("density") + 
  labs(title = "Normal Distributions")
```

```{r}
pnorm(15.5, mean = 15, sd = 0.5, lower.tail = FALSE)
```

## Beta


```{r, multBeta, fig.align="center",out.width="60%",fig.cap= '(ref:multBeta)', echo=FALSE, warning=FALSE, message=FALSE}
#betaPlots
x=seq(0,1,by=0.01)  # possible values
prob1 <- dbeta(x, 2, 5)  
prob2 <- dbeta(x, .5, .5)
prob3 <- dbeta(x, 2, 2)
prob4 <- dbeta(x, 4, 1)
betaDf <- tibble(x,prob1, prob2, prob3, prob4) %>%
  rename(x = x,
         `Beta(2, 5)` = prob1,
         `Beta(0.5, 0.5)` = prob2,
         `Beta(2, 2)` = prob3,
         `Beta(4, 1)` = prob4) %>%
  gather(2:5, key = "Distribution", value = "value") %>%
  mutate(Distribution = factor(Distribution, 
    levels = c("Beta(0.5, 0.5)", "Beta(4, 1)", "Beta(2, 2)",
               "Beta(2, 5)"))) #manual reorder

ggplot(data = betaDf, 
       aes(x = x, y = value, color = Distribution)) +
  geom_line(aes(linetype = Distribution)) +
  xlab("values") + ylab("density") + 
  labs(title="Beta Distributions") + 
  xlim(0,1) + ylim(0,4)
```

```{r}
pbeta(0.8, shape1 = 4/3, shape2 = 2, lower.tail = FALSE)
```

# Testing distributions

## Chi-squared


```{r, multChisq, fig.align="center",out.width="60%",fig.cap= '(ref:multChisq)', echo=FALSE, warning=FALSE, message=FALSE}
#chisquarePlots
x=seq(0,15,by=0.01)  # possible values
prob1 <- dchisq(x,1)  # P(Y=y)
prob2 <- dchisq(x,3)
prob3 <- dchisq(x,7)

chiDf <- tibble(x,prob1, prob2, prob3) %>%
  rename(x = x,
         `1` = prob1,
         `3` = prob2,
         `7` = prob3) %>%
  gather(2:4, key = "Degrees of Freedom", value = "value") %>%
  mutate(`Degrees of Freedom` = factor(`Degrees of Freedom`, levels = c("7", "3", "1")))

ggplot(data = chiDf, 
       aes(x = x, y = value, color = `Degrees of Freedom`)) +
  geom_line(aes(linetype = `Degrees of Freedom`)) +
  xlab("values") + ylab("density") + 
  labs(title="Chi-squared Distributions") + 
  xlim(-1,15) + ylim(0,.5)
```

```{r}
pchisq(2, 10)
```

## Student's t-Distribution


```{r, multT, fig.align="center",out.width="60%",fig.cap= '(ref:multT)', echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(-5,5, by = 0.01)
probT1 <- dt(x, 1)
probT2 <- dt(x, 2)
probT3 <- dt(x, 10)
probT4 <- dnorm(x, 0, 1)

tDf <- tibble(x, probT1, probT2, probT3, probT4) %>%
  rename(x = x,
         `1` = probT1,
         `2` = probT2,
         `10` = probT3,
         `Inf` = probT4) %>%
  gather(2:5, key = "Degrees of Freedom", value = "value") %>%
  mutate(`Degrees of Freedom` = factor(`Degrees of Freedom`, levels = c("1","2","10", "Inf")))

ggplot(data = tDf, aes(x = x, y = value, 
                       color = `Degrees of Freedom`)) +
  geom_line(aes(linetype = `Degrees of Freedom`)) +
  xlab("values") + ylab("density") + 
  labs(title="t-Distributions")
```

```{r}
pt(5, 2)
```

## F-Distribution


```{r, multF, fig.align="center",out.width="60%",fig.cap= '(ref:multF)', echo=FALSE, warning=FALSE, message=FALSE}
#fPlots
x=seq(0,7,by=0.01)  # possible values
probF1 <- df(x,1,1)  # P(Y=y)
probF2 <- df(x,4,2)
probF3 <- df(x,5,10)

fDf <- tibble(x, probF1, probF2, probF3) %>%
  rename(x = x,
         `F(1, 1)` = probF1,
         `F(4, 2)` = probF2,
         `F(5, 10)` = probF3) %>%
  gather(2:4, key = "Distribution", value = "value")

ggplot(data = fDf, aes(x = x, y = value, color = Distribution)) +
  geom_line(aes(linetype = Distribution)) +
  xlab("values") + ylab("density") + 
  labs(title="F Distributions") + 
  xlim(0,3) + ylim(0,1)
```

```{r}
pf(2, 4, 4)
```


# Table of distributions

```{r, include = FALSE}
distName <- c("Binomial","Geometric","Negative Binomial", "Hypergeometric", "Poisson", "Exponential", "Gamma", "Normal", "Beta")
cdf <- c("${n \\choose y} p^y (1-p)^{n-y}$", #binom
         "$(1-p)^yp$", #geom
         "${y + r - 1\\choose r-1} (1-p)^{y}(p)^r$", #negbinom
         "${m \\choose y}{N-m \\choose n-y}\\big/{N \\choose n}$",  #hgeom
         "${e^{-\\lambda}\\lambda^y}\\big/{y!}$", #poiss
         "$\\lambda e^{-\\lambda y}$", #exp
         "$\\displaystyle\\frac{\\lambda^r}{\\Gamma(r)} y^{r-1} e^{-\\lambda y}$", #gamma
         "$\\displaystyle\\frac{e^{-(y-\\mu)^2/ (2 \\sigma^2)}}{\\sqrt{2\\pi\\sigma^2}}$", #normal
         "$\\frac{\\Gamma(\\alpha + \\beta)}{\\Gamma(\\alpha)\\Gamma(\\beta)} y^{\\alpha-1} (1-y)^{\\beta-1}$")
params <- c("$p,\\ n$", 
            "$p$", "$p,\\ r$",
            "$n,\\ m,\\ N$",
            "$\\lambda$",
            "$\\lambda$",
            "$\\lambda, \\ r$",
            "$\\mu,\\ \\sigma$",
            "$\\alpha,\\ \\beta$")
values <- c("$0, 1, \\ldots , n$",
            "$0, 1, \\ldots, \\infty$",
            "$0, 1, \\ldots, \\infty$",
            "$0, 1, \\ldots , \\min(m,n)$",
            "$0, 1, \\ldots, \\infty$",
            "$(0, \\infty)$",
            "$(0, \\infty)$",
            "$(-\\infty,\\ \\infty)$",
            "$(0,\\ 1)$")
desc <- c("Number of successes after $n$ trials",
          "Number of failures until the first success",
          "Number of failures before $r$ successes",
          "Number of successes after $n$ trials without replacement",
          "Number of events in a fixed interval",
          "Wait time for one event in a Poisson process",
          "Wait time for $r$ events in a Poisson process",
          "Used to model many naturally occurring phenomena",
          "Useful for modeling probabilities")
```

```{r, distTable, echo = FALSE}
table1chp3 <- data.frame(distName, cdf, params, values, desc)
colnames(table1chp3) <- c("Distribution Name","pmf / pdf",
  "Parameters","Possible Y Values", "Description")
if (knitr:::is_latex_output()) {
  kable(table1chp3, booktabs = T, escape = F,
        caption="Review of mentioned random variables.") %>%
  column_spec(5, width = "10em") %>%
  kable_styling(latex_options = "scale_down")
} else {
  kable(table1chp3, "html", booktabs = T, caption="Review of mentioned random variables.")
}
```
